Legacy Data Confounds Genomics Studies

### 2019-04-30 legacy.tex
minor tweaks

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population membership as covariates. 
These models were broadly consistent (See Figure \ref{CompareModel}).

@@

\subsection{Testing the association of quality to genotype}
When conducting a statistical analysis of population genetics data, we must account for population structure. In a typical GWAS, we are interested in modelling the phenotype as a function of the genotype. 
Here we have the opposite situation, where the quantitative variable ($Q$) is used as an explanatory variable. 
So we consider models where the genotype $y$ is a function of an expected frequency $\pi_{si},$ based on population structure, and $Q$. 
The null model is 
\begin{align} \label{lfa}
y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big).
\end{align} 
The expected frequency for a SNP $s$ and individual $i$ can be estimated using principal component analysis, categorical population labels, or logistic factor analysis \citep{song2015testing}. The alternative model then takes in $Q$ as a covariate:

\begin{align}\label{gcat}
 y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg).
\end{align}

Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. 
\sgcomment{Not sure why, but the equation numbers don't line up}
$\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.

@@

To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing}.
The GCAT is a regression approach that assumes the following model.
\begin{align}\label{gcat}
 y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
\end{align} 
for $s=1\hdots S$ and $i=i,\cdots, I$  ($S = 28,516,063$ and $I = 2,504$) and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. 
The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model. 
We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs ($M =2,306,130$ from the 1kGP OMNI 2.5). 
\sgcomment{This is not consistent with what is said in the text. Also, population labels? what is M?}
To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{OMNI 2.5M chip}

>>>>

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
These models were broadly consistent (See Figure \ref{CompareModel}).

@@

\subsection{Testing the association of quality to genotype}
When conducting a statistical analysis of population genetics data, we must account for population structure. In a typical GWAS, we are interested in modelling the phenotype as a function of the genotype. 
Here we have the opposite situation, where the quantitative variable ($Q$) is used as an explanatory variable. 
So we consider models where the genotype $y$ is a function of an expected frequency $\pi_{si},$ based on population structure, and $Q$. 
The null model is 
\begin{align} \label{lfa_basic}
y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big).
\end{align} 
The expected frequency for a SNP $s$ and individual $i$ can be estimated using principal component analysis, categorical population labels, or logistic factor analysis \citep{song2015testing}. The alternative model then takes in $Q$ as a covariate:

\begin{align}\label{gcat_basic}
 y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg).
\end{align}

Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \eqref{gcat_basic} reduces to Model \eqref{lfa_basic}. 
$\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.

@@

To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing}.
The GCAT is a regression approach that assumes the following model.
\begin{align}\label{gcat}
 y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
\end{align} 
for $s=1\hdots S$ and $i=i,\cdots, I$  ($S = 28,516,063$ and $I = 2,504$) and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. 
The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model. 
We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M = 2,306,130$ SNPs (we picked SNPs from the 1kGP OMNI 2.5). 
To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{OMNI 2.5M chip}
